---
title: "Getting Started with binmetrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with binmetrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  dpi = 100
)
```

```{r setup}
library(binmetrics)
```

## Overview

**binmetrics** is built around the **CSI Framework** — a unified approach to binary
classification evaluation developed in the doctoral dissertation
*"The CSI Framework for the Evaluation of Binary Classifications in Rare Events"*
(Perales, UPNA, 2026).

The central distinction the package enforces is between two fundamentally different
types of metrics:

- **Agreement metrics** — how similar are the predictions to the observations?
- **Skill metrics** — does the classifier outperform a random baseline?

A model can have high Agreement simply because the majority class dominates. In
land change modelling, for example, a classifier that never predicts any change can
score Overall Agreement above 95% while being completely uninformative. Skill
metrics correct for this by comparing against what random guessing would achieve.

---

## 1. The contingency table

Every binary evaluation starts with a **contingency table**. Given a set of
predictions and observations, four outcomes are possible:

|                        | Observed presence | Observed absence |
|------------------------|:-----------------:|:----------------:|
| **Predicted presence** | **H** — Hits (TP) | **F** — False Alarms (FP) |
| **Predicted absence**  | **M** — Misses (FN) | **C** — Correct Rejections (TN) |

We use Finley's (1884) classic tornado forecast dataset as a running example:
28 observed tornadoes were correctly forecast (Hits), 72 were false alarms,
23 real tornadoes were missed, and 2680 non-events were correctly classified.

```{r confusion-matrix}
cm <- confusion_matrix(hits = 28, fa = 72, misses = 23, cr = 2680)
print(cm)
```

```{r confusion-summary}
summary(cm)
```

---

## 2. Agreement metrics

Agreement metrics measure the similarity between predictions and observations.
They do **not** compare against any baseline — a high Agreement value does not mean
the classifier is good.

```{r agreement-metrics}
oa(cm)          # Overall Agreement:        (H + C) / N
pa(cm)          # Producer's Accuracy (PA): H / (H + M)  =  Sensitivity
ua(cm)          # User's Accuracy (UA):     H / (H + F)  =  Precision
sp(cm)          # Specificity (SP):         C / (C + F)
far(cm)         # False Alarm Rate (FAR):   F / (C + F)  =  1 - SP
for_metric(cm)  # False Omission Rate:      M / (C + M)
csi(cm)         # Critical Success Index:   H / (H + F + M)
f1_score(cm)    # F1 Score:                 2H / (2H + F + M)
```

**CSI (Critical Success Index)** — also called Figure of Merit, Threat Score, or
Jaccard Index depending on the field — is the central metric of this framework.
It focuses exclusively on the event of interest, ignoring Correct Rejections. This
makes it appropriate when the event is rare and C dominates the table.

You can pass raw counts directly when `cr` is not needed:

```{r csi-raw}
csi(28, 72, 23)   # same result, no cr required
f1_score(28, 72, 23)
```

---

## 3. Skill metrics

Skill metrics measure how much the classifier outperforms a random baseline.
A random classifier is one that classifies presence uniformly at the rate of
prevalence — it has no actual discrimination ability.

```{r skill-metrics}
gss(cm)   # Revised Gilbert Skill Score: (CSI - CSI_random) / (1 - CSI_random)
hss(cm)   # Heidke Skill Score:          based on OA vs random OA
pss(cm)   # Peirce Skill Score:          PA - FAR
mcc(cm)   # Matthews Correlation Coeff:  Pearson corr on binary outcomes
```

**Note on GSS:** The standard Gilbert Skill Score formula has a conceptual
inconsistency — it uses the actual classifier's union (H+F+M) for the random
baseline, when the random classifier has a different union. The `gss()` function
implements the **Revised GSS** from the dissertation, which correctly uses
`CSI_random = H_rand / (H_rand + F_rand + M_rand)`.

**Note on MCC:** Although MCC is often presented as an Agreement metric in other
packages and papers, the dissertation classifies it as a **Skill metric** because
it implicitly compares against a random performance through its covariance
formulation.

**Note on Kappa:** Cohen's Kappa is mathematically equivalent to HSS for binary
classification. Its use is discouraged by the dissertation; prefer `hss()` instead.

```{r all-metrics}
# Full metric summary, grouped by Agreement then Skill.
# Returns a tibble. Skill metrics carry the baseline_agreement value (what the
# corresponding agreement metric would be for a random classifier at the same
# prediction rate) and corresponding_agreement_metric (which agreement metric
# that is: GSS -> CSI, HSS -> OA, PSS -> PA, MCC -> OA).
all_metrics(cm)
```

---

## 4. Multi-threshold analysis — TOC and CSI curves

Single-threshold metrics evaluate performance at one decision point. But a soft
classifier (one that assigns scores rather than binary labels) can be evaluated at
every possible threshold. The **TOC curve** and the **CSI curve** visualise this.

### Building a TOC data object

The `toc_data` object is the package's central data structure: a tibble with one
row per threshold `k` (the number of predicted positives).

```{r toc-from-scores}
set.seed(42)
n      <- 1000
scores <- runif(n)
obs    <- scores > 0.5 + rnorm(n, sd = 0.25)

td <- toc_from_scores(scores, obs)
head(td[, c("k", "hits", "fa", "misses", "cr", "csi", "csi_upper", "csi_baseline")])
```

Key columns:

| Column | Description |
|--------|-------------|
| `k` | Number of predicted positives (0 to N) |
| `hits`, `fa`, `misses`, `cr` | Contingency table entries at threshold `k` |
| `csi` | CSI value of the classifier at `k` |
| `csi_upper` | Upper bound — best possible CSI at `k` |
| `csi_lower` | Lower bound — worst possible CSI at `k` |
| `csi_baseline` | Random classifier baseline at `k` |
| `tpr`, `fpr` | True/False Positive Rates (for ROC) |

### TOC curve

The TOC curve plots **Hits** against `k`. The upper and lower bounds show the
range of achievable performance; the baseline represents random classification.
An optional `reference_cm` argument overlays a single-threshold forecast as a
labelled point — here Finley's 1884 tornado data.

```{r plot-toc}
cm_finley <- confusion_matrix(hits = 28, fa = 72, misses = 23, cr = 2680)
plot_toc(td, percent = TRUE,
         reference_cm = cm_finley, reference_label = "Finley (1884)")
```

### CSI curve

The CSI curve plots **CSI(k)** against `k`. It inherits the same three reference
lines. The reference point shows Finley's CSI at his operating threshold. The
x coordinate uses the reference cm's own N for normalisation, so the point is
always comparable on the percentage scale.

```{r plot-csi}
plot_csi(td, reference_cm = cm_finley, reference_label = "Finley (1884)")
```

The annotated values on the CSI plot are:

- **MaxCSI** — maximum CSI achievable at the optimal threshold
- **AUCSI** — area between the CSI curve and the lower bound, normalised
- **AUCSIS** — CSI skill score: how much AUCSI exceeds the random baseline

### ROC curve

For the ROC curve the reference point uses (FPR, TPR) — normalised rates that
are always on [0, 1]², making cross-dataset overlays straightforward. Finley's
forecast can thus be plotted on the same axes regardless of its dataset size.

```{r plot-roc}
plot_roc(td, reference_cm = cm_finley, reference_label = "Finley (1884)")
```

### Any metric across thresholds

Any metric can be plotted against `k` using `toc_metric_curve()`:

```{r plot-metric-gss}
mc_gss <- toc_metric_curve(td, gss, "GSS")
plot_metric_curve(mc_gss, "GSS")
```

```{r plot-metric-pss}
mc_pss <- toc_metric_curve(td, pss, "PSS")
plot_metric_curve(mc_pss, "PSS")
```

---

## 5. Integrated metrics

Integrated metrics summarise classifier performance **across all thresholds**.
The dissertation introduces four CSI-based integrated metrics as alternatives
to the AUC framework, designed to be less sensitive to prevalence.

```{r integrated}
integrated_metrics(td)
```

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| **MaxCSI** | max(CSI curve) | Best CSI at optimal threshold |
| **AUCSI** | ∫(CSI − CSI_lower) / ∫(CSI_upper − CSI_lower) | Normalised area under CSI curve |
| **AUCSI_baseline** | Same integral for random classifier | Chance-level AUCSI |
| **AUCSIS** | (AUCSI − AUCSI_baseline) / (1 − AUCSI_baseline) | CSI skill across all thresholds |
| **AUC** | ∫ TPR d(FPR) | Standard area under ROC curve |
| **AUCS** | 2 × (AUC − 0.5) | AUC skill score |
| **AUPRC** | ∫ Precision d(Recall) | Area under Precision-Recall curve |

The key advantage of AUCSI and AUCSIS over AUC and AUCS is that they exclude
Correct Rejections, focusing exclusively on the rare event. This makes them less
inflated when prevalence is low.

---

## 6. Object-oriented API — `BinaryClassifier`

The `BinaryClassifier` R6 class wraps the full workflow into a single object.

```{r oo-api}
clf <- BinaryClassifier$new(scores, obs, name = "Demo Classifier")
clf
```

```{r oo-summary}
clf$summary()
```

All integrated metrics are available as active bindings:

```{r oo-bindings}
clf$mcsi       # MaxCSI
clf$aucsi      # AUCSI
clf$aucsis     # AUCSIS
clf$auc        # AUC
clf$aucs       # AUCS
clf$auprc      # AUPRC
```

Evaluate any threshold directly:

```{r oo-threshold}
cm_at_200 <- clf$at_threshold(200)
csi(cm_at_200)
gss(cm_at_200)
```

### Plots from the object

```{r oo-plots, eval=FALSE}
clf$plot_toc()
clf$plot_csi()          # CSI curve with AUCSI / AUCSIS / MaxCSI annotations
clf$plot_roc()
clf$plot_all_metrics()  # 2×4 panel: Agreement (OA/CSI/PA/SP) + Skill (GSS/HSS/PSS/MCC)
```

```{r oo-plots-shown}
clf$plot_all_metrics()
```

---

## 7. Building from pre-computed hit counts

When you already have cumulative hit counts (e.g. from a published TOC table),
use `BinaryClassifier_from_hits()`:

```{r from-hits}
P <- 5; Q <- 10
# Perfect classifier: all positives ranked before any negative
hits_perfect <- c(0, 1, 2, 3, 4, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5)
clf_perfect  <- BinaryClassifier_from_hits(hits_perfect, P, Q, name = "Perfect")
clf_perfect$mcsi   # 1
clf_perfect$auc    # 1
clf_perfect$aucsi  # 1
```

---

## 8. Comparing classifiers

`compare_classifiers()` overlays multiple classifiers on the same plot.

```{r compare}
set.seed(1)
td_a <- toc_from_scores(runif(500), runif(500) > 0.4)
td_b <- toc_from_scores(runif(500) ^ 0.5, runif(500) > 0.4)

compare_classifiers(list("Random" = td_a, "Better" = td_b), type = "csi")
compare_classifiers(list("Random" = td_a, "Better" = td_b), type = "roc")
```

---

## 9. Interactive apps

Two Shiny applications allow interactive exploration:

```{r shiny, eval=FALSE}
# Slide the threshold k and watch all metrics update in real time
launch_threshold_explorer(td)

# Explore how prevalence shapes the feasible region in AUC × AUCSI space
launch_bounds_explorer()
```

---

## Summary

The CSI Framework provided by **binmetrics** rests on three pillars:

1. **Distinguish Agreement from Skill.** Agreement tells you how similar predictions
   are to observations; Skill tells you whether the classifier is actually better than
   random. These are different questions requiring different metrics.

2. **Focus on the rare event.** CSI, AUCSI, and AUCSIS ignore Correct Rejections,
   making them robust to class imbalance. OA and AUC inflate under low prevalence.

3. **Analyse the full threshold range.** A single operating point hides information
   about the soft classifier's ranking ability. The TOC and CSI curves — together
   with their bounds and baseline — expose Quantity and Allocation disagreement
   at every threshold.
