---
title: "Getting Started with binmetrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with binmetrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  dpi = 100
)
```

```{r setup}
library(binmetrics)
```

## Overview

**binmetrics** evaluates binary classifiers through the full range of decision
thresholds, not just at a single operating point. The package centres on the
**Threshold Operating Characteristic (TOC)** space, which represents classifier
performance as a function of the number of predicted positives $k$.

Two families of metrics are distinguished:

- **Agreement metrics** (OA, BA, MCC, Kappa, F1): compare predictions to
  observations with no notion of a "baseline".
- **Skill metrics** (FOM/CSI, GSS, PSS, HSS): measure improvement over a
  random or uninformative classifier.

---

## 1. Single-threshold analysis

The simplest entry point is a confusion matrix at a fixed threshold.

```{r single-threshold}
# Finley (1884) tornado forecasting data
cm <- confusion_matrix(hits = 28, fa = 72, misses = 23, cr = 2680)
print(cm)
```

```{r single-threshold-summary}
summary(cm)
```

You can call any metric function directly:

```{r single-metrics}
fom(cm)           # Figure of Merit / CSI
gss(cm)           # Gilbert Skill Score (= ETS)
pss(cm)           # Peirce Skill Score
all_metrics(cm)   # all at once
```

---

## 2. Full threshold analysis — TOC data

The key object in **binmetrics** is the `toc_data` tibble, which enumerates
the classifier's performance at every possible threshold rank.

### From classifier scores

```{r toc-from-scores}
set.seed(42)
n      <- 1000
scores <- runif(n)
obs    <- scores > 0.5 + rnorm(n, sd = 0.25)

td <- toc_from_scores(scores, obs)
td
```

### TOC plot

```{r plot-toc}
plot_toc(td)
```

### ROC plot

```{r plot-roc}
plot_roc(td)
```

### FOM curve with theoretical bounds

The FOM curve is shown alongside the **upper bound** (best possible classifier
with the same hit counts), the **lower bound** (worst-case ranking), and the
**random classifier** baseline.

```{r plot-fom}
plot_fom(td)
```

The annotated values are:

- **AFOM**: area under FOM curve normalised by the upper–lower band area.
- **DFOM**: FOM skill — improvement of AFOM over the random baseline.
- **MFOM**: maximum FOM across all thresholds.

### Any metric with its bounds

```{r plot-metric}
# GSS curve with bounds
mc <- toc_metric_curve(td, gss, "GSS")
plot_metric_curve(mc, "GSS")
```

---

## 3. Integrated metrics

```{r integrated}
integrated_metrics(td)
```

| Metric | Meaning |
|--------|---------|
| MFOM   | Maximum FOM across all thresholds |
| AFOM   | Area-normalised integrated FOM |
| AUFOM  | AFOM of a random classifier (baseline) |
| DFOM   | FOM skill = (AFOM - AUFOM) / (1 - AUFOM) |
| AUC    | Area under ROC curve |
| DAUC   | AUC skill = 2*(AUC - 0.5) |
| AUCPRC | Area under Precision-Recall curve |

---

## 4. Object-oriented API

The `BinaryClassifier` R6 class bundles the full workflow.

```{r oo-api}
clf <- BinaryClassifier$new(scores, obs, name = "Demo Classifier")
clf$summary()
```

```{r oo-plots, eval=FALSE}
clf$plot_toc()
clf$plot_fom()
clf$plot_roc()
clf$plot_all_metrics()   # patchwork panel of all 8 standard metrics
```

---

## 5. Comparing classifiers

```{r compare}
set.seed(1)
clf_a <- toc_from_scores(runif(500), runif(500) > 0.4)
clf_b <- toc_from_scores(runif(500), runif(500) > 0.5)

compare_classifiers(list(A = clf_a, B = clf_b), type = "roc")
```

---

## 6. Interactive apps

```{r shiny, eval=FALSE}
# Threshold slider: move k and watch metrics update live
launch_threshold_explorer(td)

# Explore the feasible region of AUC and AFOM for different prevalences
launch_bounds_explorer()
```
